Effects of resonant and random excitations on the proton beam in the Large Hadron Collider, with applications to the design of pulsed hollow electron lenses for active halo control

### 2018-04-19 resex.tex
Implemented Miriam's comments on circulated draft

\begin{table*}
  \caption{ Stored beam energy for different past, present and future
    colliders. Each new machine represents a leap in stored beam
    energy.}
  \label{tab:stored_energy}
  \begin{ruledtabular}
    \begin{tabular}{lccccc}
      Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{Wenninger:IPAC:2017}	& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
      \colrule
      Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
      Number of bunches & 36 & 2220 & 2808 & 2748 & ? \\
      Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
      Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

\begin{figure*}
  \begin{tabular}{cc}
    No excitation & 10th, H+V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    7th, H & 7th, V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); \tenthtp\ in both
    horizontal and vertical planes (top left); \seventhtp\ in the
    horizontal (bottom left) and vertical (bottom right) planes. The
    excitation amplitude is 120~nrad in the corresponding plane. The
    colors (blue to red) represent the tune jitter of tracked
    particles starting at each given location in tune
    space~\cite{fmalaskar}. The absence of a strong excitation of any
    resonance in case of vertical \seventhtp\ and the strong
    excitation in case of horizontal pulsing confirms the excitation
    of the $7 Q_x$ resonance. For \tenthtp, there is in contrast no
    significant difference between H, V, or H+V
    (Fig.~\ref{fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

@@

The effects of 7th- and \tenthtp\ are also observed in the absence of
machine lattice errors. In this case, the only sources of nonlinearity
are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer},
suggesting that these nonlinearities are responsible for the
pronounced beam sensitivity. The driven resonances are revealed by the
frequency-map analysis, shown in Fig.~\ref{fig:patternfma}. The
$7 Q_x$ resonance is excited by the \seventhtp, and the $10 Q_x$ and
$10 Q_y$ resonances by the \tenthtp. As octupoles can only drive even
resonances, the sources of the $7 Q_x$ resonances are the sextupoles,
while the octupoles generate the tune footprint. The other pulsing
patterns do not exhibit any increase in losses or emittance growth
without magnetic errors. Their effect can be thus attributed to an
interaction of the excitation with the lattice errors, implying also a
sensitivity to the chosen random seed in simulations and to the
specific lattice in experiments.

@@

Although on a different time scale (minutes instead of seconds), a
similar re-adjustment of the beam distribution in response to an
external excitation was directly observed experimentally for the first
time, to our knowledge, during the 2016 and 2017 LHC experiments
presented in this paper, as discussed below.

If only the resonant excitation is present, without noise, simulations
show that the modified distribution is stable, with constant
emittance. By adding a random noise component, representative of the
natural noise present in the LHC, one calculates, after the initial
adjustment phase, emittance-growth rates that increase with the
amplitude of the excitation. As discussed below, an increase of the
emittance-growth rates with the applied excitation amplitude is what
is observed experimentally.

@@

A direct comparison of measured and predicted loss rates is presented
in Fig.~\ref{fig:10thexploss}. As experiments and simulations were
conducted on different time intervals (11~minutes vs.\ 90~s), we
compare the relative average loss rate~$R$:
%
\begin{equation}
  \label{eqn:lossrate}
  R = \frac{I_{\mathrm{start}} - I_{\mathrm{end}}}{I_{\mathrm{start}}}
  \cdot \frac{1}{\Delta t},
\end{equation}
%
where $I$ is the beam intensity and
$\Delta t = t_\mathrm{end} - t_\mathrm{start}$ is the time interval
during which the excitation with given amplitude is
applied. Uncertainties on the excitation amplitude arise from the
calibration of the transverse damping system
(Section~\ref{sec:adt}). For both experiments and simulations, the
error bars on the loss rate are statistical. The magnitude of the
random noise component introduces a systematic uncertainty in the
calculations, which was not evaluated for the scope of this paper. The
systematic uncertainty on the measurements can be inferred from the
difference between the 2~periods of the experiment, which should in
principle be the same. Part of this systematic effect could be due to
the fact that the second set of excitations with double amplitude
started with a modified beam distribution. The magnitude of this
effect could be verified in future experiments in which the
excitations are reversed or where a new beam fill is used for each set
of measurements. Overall, we conclude that the effects of this
resonant excitation could be measured in the LHC within a factor~2 and
could be predicted with similar accuracy.

@@

The machine tunes were changed in standard operations from (64.28,
59.31) in~2016 to (62.27, 60.295) in~2017. This change was accompanied
by a small change in optics, considered negligible for these
measurements. A direct comparison of the two experiments is therefore
not possible. However, these differences provide an estimate of how
the effects of resonant excitations are affected by lattice changes.

@@

The effects on the beam were tested during the 2017 experiment. The
excitation amplitude was increased to a maximum of 96~nrad. The loss
rates as a function of excitation amplitude are presented in
Fig.~\ref{fig:8thexploss2017}. For H and V excitations, no significant
increase of the loss rate was observed. A weak increase with amplitude
was seen in the case of H+V excitation. Observed and predicted loss
rates are of the same magnitude (about 0.2\%/h).

Changes in beam distributions were also detected
(Fig.~\ref{fig:8thexpprof}). The horizontal distribution responded to
the H excitation with a slight depletion of the core and a
corresponding increase around $2\sigma$. For an H+V excitation, one
would expect a similar distribution change. However, this change could
not be detected directly, because, in this case, the emittance of the
control bunches was much smaller (\q{1.8}{\mu m}) than the emittance
of the excited bunches (\q{2.6}{\mu m}), and intrabeam scattering
masked the effects of the excitation~\cite{resexmd2017}. In the
vertical plane, beam distributions and emittances were not affected
by the resonant excitation.

@@

The effects of random excitation were measured during the 2017
experiments. Increasing emittance growth with excitation amplitude was
recorded only in the plane of excitation. Horizontal and vertical
emittance growth rates for a given amplitude were comparable. As an
example, the relative horizontal and vertical emittances during V and
H+V pulsing are illustrated in Fig.~\ref{fig:ranexp}. In addition, the
excitations translated into constant growth rates. In particular,
there was no adjustment phase followed by an equilibrium state, as
observed during 7th- and \tenthtp. The qualitative behavior of
emittances predicted in simulations was therefore confirmed. However,
the calculations could only roughly reproduce the magnitude of the
measured effects and were in general lower. For instance, at 6~nrad,
the calculated emittance growth was 1\%/min, whereas in experiments we
measured 5\%/min.

@@

The effects of resonant excitations depend on several factors,
including the machine lattice and the distribution of the beam in
phase space and in betatron tunes. For this reason, we decided to
study resonant transverse kicks on proton beams directly in the
LHC. We relied on the flexibility of the LHC transverse feedback and
damping system (ADT) to emulate the nonlinear residual kicks of a
hollow electron lens with transverse dipole kicks, using various
excitation patterns. We focused on the effects on the beam core,
namely losses, emittances and beam distributions, as these determine
the magnitude of the residual fields that can be tolerated. Obviously,
the choice of pulsing patterns to be used in operations depends also
on their effects on the beam halo. These effects are outside the scope
of the present work (and they are challenging to measure directly).

@@

The most effective pulsing patterns were identified by numerical
tracking simulations and by frequency-map analysis. These calculations
provided insights on the rich nonlinear dynamics of resonant
excitations. Kicking the beam every 7th turn and every 10th turn were
chosen as examples of strong excitations, whereas \eighthtp\ had very
weak consequences.

@@

The experiments confirmed the relative strength of the pulsing
modes. For instance, random and \seventhtp\ generated losses at the
level of a few percent per hour at 12~nrad, whereas the same level of
losses was obtained at 96~nrad for \tenthtp; even at 96~nrad, losses
from \eighthtp\ did not exceed 0.6\%/h. A similar hierarchy was
observed with respect to emittance growth. In many cases, simulations
underestimated losses and emittance growth. One important unknown
factor was the magnitude of machine noise and its sources, which had
to be extrapolated from collision to injection energy. The comparison
between \seventhtp\ in 2016 and 2017 emphasized the sensitivity of
horizontal and vertical excitations to working point and tune
footprint.

>>>>

\begin{table*}
  \caption{ Stored beam energy for different past, present and future
    colliders. Each new machine represents a leap in stored beam
    energy.}
  \label{tab:stored_energy}
  \begin{ruledtabular}
    \begin{tabular}{lccccc}
      Collider& Tevatron (protons) \cite{tevatron} & LHC 2016 \cite{Wenninger:IPAC:2017}	& LHC nominal \cite{lhc_design} & HL-LHC \cite{hlcdr} & FCC \cite{fcc_param_2017} \\
      \colrule
      Beam energy [TeV] & 0.98 & 6.5 & 7.0 & 7.0 & 50.0\\
      Number of bunches & 36 & 2220 & 2808 & 2748 & $\sim$10500 \\
      Number of particles per bunch & $2.90\times 10^{11}$ & $1.15\times 10^{11}$ & $1.15\times 10^{11}$ & $2.2\times 10^{11}$ & $1.0\times 10^{11}$\\
      Stored beam energy [MJ] & 1.6 & 265.9 & 362.2 & 678.0 & 8400 \\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

@@

\begin{figure*}
  \begin{tabular}{cc}
    No excitation & 10th, H+V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    7th, H & 7th, V \\
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=\fmawidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); \tenthtp\ in both
    horizontal and vertical planes (top right); \seventhtp\ in the
    horizontal (bottom left) and vertical (bottom right) planes. The
    excitation amplitude is 120~nrad in the corresponding plane. The
    colors (blue to red) represent the tune jitter of tracked
    particles starting at each given location in tune
    space~\cite{fmalaskar}. The absence of a strong excitation of any
    resonance in case of vertical \seventhtp\ and the strong
    excitation in case of horizontal pulsing confirms the excitation
    of the $7 Q_x$ resonance. For \tenthtp, there is in contrast no
    significant difference between H, V, or H+V
    (Fig.~\ref{fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

@@

The effects of 7th- and \tenthtp\ are also observed in the absence of
machine lattice errors. In this case, the only sources of nonlinearity
are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer},
suggesting that these nonlinearities are responsible for the
pronounced beam sensitivity. The driven resonances are revealed by the
frequency-map analysis, shown in Fig.~\ref{fig:patternfma}. The
$7 Q_x$ resonance is excited by the \seventhtp, and the $10 Q_x$ and
$10 Q_y$ resonances by the \tenthtp. As octupoles can only drive even
resonances, the sources of the $7 Q_x$ resonances are the sextupoles,
while the octupoles generate the tune footprint. The other pulsing
patterns do not exhibit any increase in losses or emittance growth
without magnetic errors. Their effect can be thus attributed to an
interaction of the excitation with the machine errors, implying also a
sensitivity to the chosen random seed in simulations and to the
specific error distribution in experiments.

@@

Although on a different time scale (minutes instead of seconds), a
similar re-adjustment of the beam distribution was directly observed
for the first time, to our knowledge, during the 2016 and 2017 LHC
experiments presented in this paper, as discussed below.

If only the resonant excitation is present, without noise, simulations
show that the modified distribution is stable, with constant
emittance. By adding a random noise component, representative of the
natural noise present in the LHC, one calculates, after the initial
adjustment phase, emittance-growth rates that increase with the
amplitude of the excitation. As discussed below, an increase of the
emittance-growth rates with the applied excitation amplitude was also
observed experimentally.

@@

A direct comparison of measured and predicted loss rates is presented
in Fig.~\ref{fig:10thexploss}. As experiments and simulations were
conducted on different time intervals (11~minutes vs.\ 90~s), we
compare the relative average loss rate~$R$:
%
\begin{equation}
  \label{eqn:lossrate}
  R = \frac{I_{\mathrm{start}} - I_{\mathrm{end}}}{I_{\mathrm{start}}}
  \cdot \frac{1}{\Delta t},
\end{equation}
%
where $I$ is the beam intensity and
$\Delta t = t_\mathrm{end} - t_\mathrm{start}$ is the time interval
during which the excitation with given amplitude is
applied. Uncertainties on the excitation amplitude arise from the
calibration of the transverse damping system
(Section~\ref{sec:adt}). For both experiments and simulations, the
error bars on the loss rate are statistical. The magnitude of the
random noise component introduces a systematic uncertainty in the
calculations, which was not evaluated for the scope of this paper. The
systematic uncertainty on the measurements can be inferred from the
difference between the 2~periods of the experiment, which should in
principle be the same. Part of this systematic effect could be due to
the fact that the second set of excitations with double amplitude
started with a modified beam distribution. The magnitude of this
effect could be verified in future experiments in which the
excitations are reversed or where a new beam fill is used for each set
of measurements. Overall, we conclude that the effects of this
resonant excitation on loss rates could be measured in the LHC within
a factor~2 and could be predicted with similar accuracy.

@@

The machine tunes were changed in standard operations from (64.28,
59.31) in~2016 to (62.27, 60.295) in~2017. This change was accompanied
by a small change in optics, considered negligible for these
measurements. A direct comparison of the two experiments is therefore
not possible. However, these differences show how the effects of
resonant excitations are affected by small changes in fractional tune.

@@

The effects on the beam were tested during the 2017 experiment. The
excitation amplitude was increased to a maximum of 96~nrad. The loss
rates as a function of excitation amplitude are presented in
Fig.~\ref{fig:8thexploss2017}. For H and V excitations, no significant
increase of the loss rate was observed. A weak increase with amplitude
was seen in the case of H+V excitation. Observed and predicted loss
rates were roughly of the same magnitude (about 0.1--0.4\%/h).

Changes in beam distributions were also detected
(Fig.~\ref{fig:8thexpprof}). The horizontal distribution responded to
the H excitation with a slight depletion of the core and a
corresponding increase around $2\sigma$. For an H+V excitation, one
would expect a similar distribution change. However, this change could
not be detected directly, because, in this case, the emittance of the
control bunches was much smaller (\q{1.8}{\mu m}) than the emittance
of the excited bunches (\q{2.6}{\mu m}), and intra-beam scattering
masked the effects of the excitation~\cite{resexmd2017}. In the
vertical plane, beam distributions and emittances were not affected
by the resonant excitation.

@@

The effects of random excitation were measured during the 2017
experiments. Increasing emittance growth with excitation amplitude was
recorded only in the plane of excitation. Horizontal and vertical
emittance growth rates for a given amplitude were comparable. As an
example, the relative horizontal and vertical emittances during V and
H+V pulsing are illustrated in Fig.~\ref{fig:ranexp}. In addition, the
excitations translated into constant growth rates. In particular,
there was no adjustment phase followed by an equilibrium state, as
observed during 7th- and \tenthtp. The qualitative behavior of
emittances predicted in simulations was therefore confirmed.
Calculations could also reproduce the magnitude of the measured
effects. For instance, at 6~nrad, the calculated emittance growth
rates were 3.6\%/min in the horizontal plane and 2.7\%/min vertically,
whereas in experiments we measured 5.0\%/min.

@@

The effects of resonant excitations depend on several factors,
including machine optics and the distribution of the beam in phase
space and in betatron tunes. For this reason, we decided to study
resonant transverse kicks on proton beams directly in the LHC. We
relied on the flexibility of the LHC transverse feedback and damping
system (ADT) to emulate the nonlinear residual kicks of a hollow
electron lens with transverse dipole kicks, using various excitation
patterns. We focused on the effects on the beam core, namely losses,
emittances and beam distributions, as these determine the magnitude of
the residual fields that can be tolerated. Obviously, the choice of
pulsing patterns to be used in operations depends also on their
effects on the beam halo. These effects are outside the scope of the
present work (and they are challenging to measure directly).

@@

The effects of the pulsing patterns were evaluated by numerical
tracking simulations and by frequency-map analysis. These calculations
provided insights on the rich nonlinear dynamics of resonant
excitations. Kicking the beam every 7th turn and every 10th turn were
chosen as examples of strong excitations, whereas \eighthtp\ had very
weak consequences.

@@

The experiments confirmed the relative strength of the pulsing
modes. For instance, random and \seventhtp\ generated losses at the
level of a few percent per hour at 12~nrad, whereas the same level of
losses was obtained at 96~nrad for \tenthtp; even at 96~nrad, losses
from \eighthtp\ did not exceed 0.6\%/h. A similar hierarchy was
observed with respect to emittance growth; in this case, \tenthtp\ was
stronger than \seventhtp. In many cases, simulations underestimated
losses and emittance growth. One important unknown factor was the
magnitude of machine noise and its sources, which had to be
extrapolated from collision to injection energy. Other factors were
collective effects such as intra-beam scattering and electron cloud,
which were minimized in experiments but could not be entirely
avoided. The comparison between \seventhtp\ in 2016 and 2017
emphasized the sensitivity of horizontal and vertical excitations to
working point and tune footprint.
